---
title: "A multicentre observational study of the epidemiology and initial management pathway of thyroid nodules"
author:
  - name: Ovie Edafe
    orcid: 0000-0002-6205-806X
    corresponding: true
    email: ovieedafe@hotmail.co.uk
    roles:
      - Researcher
    affiliations:
      - Department of Oncology & Metabolism, University of Sheffield
  - name: Neil Shephard
    orcid: 000-0001-8301-6857
    corresponding: false
    email:
    roles:
      - Researcher
    affiliations:
      - Research Software Engineer, Department of Computer Science, University of Sheffield
  - name: Sabapathy P Balasubramanian
    orcid: 0000-0001-5953-2843
    corresponding: false
    roles: Researcher
    affiliations:
      - Directorate of General Surgery, Sheffield Teaching Hospitals NHS Foundation Trust
  - name: Thy3000 Collaborators
    orcid:
    corresponding: false
    roles: Collaborators
    affiliations:
abstract: |
  ### Introduction {.unnumbered}

  The incidence, modes of presentation, and distribution of pathology of thyroid nodules in the United Kingdom (UK) is
  not well characterised. The aim of the study is to understand the epidemiology and management pathways of thyroid
  nodules in the UK.

  ### Methods {.unnumbered}

  This was a retrospective study over a 6-month period. We included all consecutive adult patients referred for
  investigation of a thyroid nodule.

  ### Results {.unnumbered}

  A total of 12 centres including 573 patients with median age of 54 years were included. Incidental detection of
  thyroid nodule was reported in 31.6%. The U classifications were: U1 (0.5%), U2 (45.6%), U3 (44.4%), U4 (7.2%), U5
  (2.3%). Of the patients who had FNAC, the Thy classifications were: Thy1 (21.9%), Thy2 (37.5%), Thy3a (10.3%), Thy3f
  (20.9%), Thy4 (4.7%), Thy5 (4.7%). Core biopsy was performed in 8 (1.4%) patients. The initial management was
  thyroidectomy in 23.6% and conservative management in 72.8%. In the latter group, 55.8% were offered routine review in
  a median period of 24 weeks and 15% had a change in the initial management strategy. The postoperative pathology was
  malignant in 41.3% of patients. Age, BMI, previous neck irradiation, incidental nodule, and rapid growth factors were
  not significantly associated with thyroid cancer. The proportion of cancer in ‘Thy3a’ and ‘Thy3f’ nodules were 42.9%
  and 47.9% respectively.

  ### Conclusion {.unnumbered}

  Around 60% of thyroid nodules subjected to diagnostic thyroidectomy were benign. Better preoperative predictors and
  tools to differentiate between benign and malignat thyroid nodules are needed.
keywords:
  - Thyroid nodules
  - Thyroid cancer
plain-language-summary: |
    This was a multicentre study to analyse the presentation of thyroid nodules in the United Kingdom (UK). A total of 573 patients with thyroid nodules were included across 12 sites. Around 20% of patients underwent diagnostic thyroidectomy.Of the patients who underwent diagnostic thyroidectomy, around 60% had a benign pathology. This highlights the need for better preoperative factors and tools to differentiate between benign and malignant thyroid nodules.
key-points:
  - A high proportion of diagnostic thyroidectomy shows benign pathology
  - Better discrimators between benign and malignant thyroid nodules are required in the UK
date: 2024-03-21
bibliography: references.bib
citation:
  container-title: "Multicentre study on thyroid nodules"
number-sections: true
---

```{r}
#| label: setup
#| eval: true
#| echo: false
#| warning: false

# Load libraries
library(ggplot2)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(gtsummary)
library(tableone)

# Set global options
options(digits = 3)

## Set directories based on current location
base_dir <- getwd()
data_dir <- paste(base_dir, "data", sep = "/")
csv_dir <- paste(data_dir, "csv", sep = "/")
r_dir <- paste(data_dir, "r", sep = "/")
r_scripts <- paste(base_dir, "r", sep = "/")


# Load data
source("r/clean.R")
#df <- readRDS(paste0(r_dir, "clean.rds", sep="/"))
```

## Introduction

The epidemiology of thyroid nodules and in particular modes of presentation and pathology is not well characterised in
the United Kingdom (UK) @abdellateif2020.  Common reasons for referral of patients with thyroid nodules to secondary
care include neck lump, abnormal thyroid function test and compressive symptoms.  Incidental thyroid lesions are
increasingly being identified @mitchell2005. Given how common thyroid nodules are and the low yield thyroid cancer
(4-6.5%) @tamhane2016 in patients with presenting with thyroid nodules, most patients referred via this pathway are
likely to have benign disease @perros2014. The standard practice for evaluating thyroid nodules in the UK includes
ultrasound evaluation to grade the nodule as per the BTA ultrasound classification system. Patients considered to meet
threshold for FNAC are those with U3, U4 or U5 nodules @perros2014. FNAC is generally not advised if nodule < 1cm in
diameter unless there is clinical suspicious lymph node or extrathyroidal extension @mitchell2016,
@perros2014. Anecdotally, there is thought to be variability with regards to the initial management of nodules and
follow up of patients. The aim of this study was to understand the modes of presentation and distribution of pathologies
in thyroid nodules, and the  management pathways.

## Methods

### Centre eligibility

All secondary care centres investigating and managing patients with thyroid nodules were eligible for
inclusion. Potential individual principal investigators were invited to participate in this study through the British
Association of Endocrine and Thyroid surgeons (BAETS). Each centre was required to have a consultant priniciple
investigator (PI) and a trainee lead to manage the study. Centres were required to contribute consecutive patients
presenting with thyroid nodules within the study period to be eligible for participation. The target was that each site
would contribute at least 30 consecutive patients to the study. A total of 42 sites initially showed interest in
participating in the study. The potential consultant PIs of each site were contacted via email with the study protocol
and guidance for registration of the study locally. In addition, the local research offices were contacted. At least
four reminder emails were sent to each site over a six-month period. Seventeen sites completed the
local registration process, but only 14 sites contributed data towards the study. Three sites were sent reminders at
least three times over a six-month period, but no data was provided. Two sites contributed a total of 4 patients each;
they were sent reminders at least three times over a six-month period to provide more data However, no further data was
provide; these sites were excluded. Data from 12 sites were included in the final analysis.
All secondary care centres investigating and managing patients with thyroid nodules were eligible for
inclusion. Potential individual principal investigators were invited to participate in this study through the British
Association of Endocrine and Thyroid surgeons (BAETS). Each centre was required to have a consultant priniciple
investigator (PI) and a trainee lead to manage the study. Centres were required to contribute consecutive patients
presenting with thyroid nodules within the study period to be eligible for participation. The target was that each site
would contribute at least 30 consecutive patients to the study. A total of 42 sites initially showed interest in
participating in the study. The potential consultant PIs of each site were contacted via email with the
study protocol and guidance for registration of the study locally. In addition, the local research offices were
contacted. At least four reminder emails were sent to each site over a six-month period. Seventeen sites completed the
local registration process, but only 14 sites contributed data towards the study. Three sites were sent reminders at
least three times over a six-month period, but no data was provided. Two sites contributed a total of 4 patients each;
they were sent reminders at least three times over a six-month period to provide more data However, no further data was
provide; these sites were excluded. Data from 12 sites were included in the final analysis.

### Patient inclusion and exclusion criteria Inclusion criteria

All adult patients (aged 18 or above) being referred for investigation of thyroid nodule to secondary care. The
exclusion criteria were as follows: patients attending clinics for follow up appointments after investigation or
treatment or for continuation of conservative treatment; new patients referred for specialist treatment from another
secondary care setting; and patients referred again after being seen or treated previously for a thyroid nodule in the
last two years.

### Participant identification and recruitment

There was no direct contact with patients given the retrospective nature of the study. Patient records that met the
above inclusion criteria were eligible for inclusion.

### Variables

Data on the following variables were collected: patient demographics, referral details, clinical examination findings,
initial investigations, initial management, pathology results. Data was collected from existing patient clinical
records. Missing data was be noted. Data was collected using a standardised data collection form on the REDCap
database. The REDCap database is an online data collection tool established for the use in multicentre research studies.

### Data analysis

All data collected was analysed using the R Statistical Software @r_citation and the Tidyverse packages @tidyverse. An
initial descriptive analysis of the variables were performed. Continuous data was summarised as mean and standard
deviation (SD) if normally distributed, and median and interquartile range (IQR) where non-normally
distributed. Categorical data was summarised as frequency and percentages. The Chi-squared test was used to examine the
relationship between categorical variables and outcome of thyroid cancer. Independent sample t-test or the Mann-Whitney
U test was used to evaluate the association between continuous variables and thyroid cancer depending on whether the
data was normally distributed or not. We defined benign pathology as Thy2 cytology or histological confirmation of benign
pathology. Malignant nodules were confirmed on histology.

### Ethical consideration

Ethical approval was obtained from the health research authority (HRA) through the integrated research application
system (IRAS) to undertake this study (19/HRA/5351).


```{r}
#| label: counting
#| eval: true
#| echo: false
#| warning: false
n_obs <- nrow(df)
n_centre <- df |>
  select(data_access_group) |>
  distinct() |>
  nrow()
## Table centers, exclude if no recruitments
by_centre <- df |>
  select(data_access_group) |>
  droplevels() |>     ## This omits levels that are unobserved (i.e. "NHS Dumfries and Galloway" and "Wirral")
  table()


## Calculate numbers by centre, there are two ways of doing this, we can make a table and then select rows from it
clinic_recruiting = df$clinic_recruiting |> table()

## Or we can calculate the values, storing them and also using them to calculate the proportion(/percentage)
n_ent <- df |>
  filter(clinic_recruiting == "ENT") |>
  nrow()
prop_ent = 100 * n_ent / n_obs

n_gen_surg <- df |>
  filter(clinic_recruiting == "General surgery (including endocrine surgery)") |>
  nrow()
prop_gen_surg = 100 * n_gen_surg / n_obs

n_joint_thyroid <- df |>
  filter(clinic_recruiting == "Joint thyroid") |>
  nrow()
prop_joint_thy = 100 * n_joint_thyroid/n_obs

n_medicine <- df |>
  filter(clinic_recruiting == "Medicine") |>
  nrow()
prop_medicine = 100 * n_medicine / n_obs

n_other <- df |>
  filter(clinic_recruiting == "Other") |>
  nrow()
prop_other = 100 * n_other / n_obs

n_ref_gp <- df |>
  filter(referral_source == "GP") |>
  nrow()
prop_gp = 100 * n_ref_gp / n_obs

n_ref_sec <- df |>
  filter(referral_source == "Secondary") |>
  nrow()
prop_sec = 100 * n_ref_sec / n_obs

n_2ww <- df |>
  filter(two_week_wait_referral == "Yes") |>
  nrow()
prop_2ww = 100 * n_2ww / n_obs

n_current_smoker <- df |>
  filter(smoking_status == "Current smoker") |>
  nrow()
prop_current_smoker = 100 * n_2ww / n_obs

n_ex_smoker <- df |>
  filter(smoking_status == "Ex smoker > 6 weeks") |>
  nrow()
prop_ex_smoker = 100 * n_ex_smoker / n_obs

n_non_smoker <- df |>
  filter(smoking_status == "Non-smoker") |>
  nrow()
prop_non_smoker = 100 * n_non_smoker / n_obs

bmi_recoded <- nrow(df$bmi)

sites <- df %>% select(data_access_group) %>% tbl_summary()

referral <- df %>% select(clinic_recruiting, referral_source, two_week_wait_referral) %>% tbl_summary()
```



## Results

A total of `r n_centre` centres were included in this study, with between `r min(by_centre)` and `r max(by_centre)`
patients, contributing a total of `r n_obs` patients (see @tbl-recruitment). The mean (SD) of patients include
was `r mean(df$age, na.rm=TRUE)` (`r sd(df$age, na.rm=TRUE)`) 
<!--(see @tbl-continuous_summary).  Patients were identified
from a range of clinics with the majority coming from Ear, Nose and Throat (ENT) (n=`r clinic_recruiting["ENT"]`; `r
100 * (clinic_recruiting["ENT"] / n_obs) `%). The second most common clinic was General Surgery (n=`r
clinic_recruiting["General surgery (including endocrine surgery)"]`;`r 100 * (clinic_recruiting["General surgery
(including endocrine surgery)"] / n_obs)`%). See table for the distribution of clinics @tbl-clinics. Sources of referral
for thyroid nodules were GP and Secondary care in 423 (74.0%) and 150 (26.0%) respectively. Nodules were referred
through the NHS as two-week wait (2WW) cancer pathway in 52.4% of cases-->


```{r}
#| label: tbl-recruitment
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Site recruitment"
df |>
  select(data_access_group) |>
  gtsummary::tbl_summary() |>
  gtsummary::modify_caption("Site Recruitment")
```

```{r}
#| label: tbl-clinics
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Referring Clinics"
df |>
select(clinic_recruiting) |>
  gtsummary::tbl_summary() |>
  gtsummary::modify_caption("Referring Clinics")
```


```{r}
#| label: tbl-scans
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Scans"
df |>
  select(
  ultrasound,
  iodine_scan,
  ct_neck,
  mri_neck) |>
  gtsummary::tbl_summary() |>
  gtsummary::modify_caption("Scans")
```

```{r}
#| label: tbl-patient-characteristics-gtsummary
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Previous neck irradiation"
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
patient_characteristics <- df |>
  dplyr::ungroup() |>
  dplyr::select(c(
    "age", "bmi", "referral_source", "two_week_wait_referral", "smoking_status",
    "previous_neck_irradiation", "asa_score"
  )) |>
gtsummary::tbl_summary(
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{N_nonmiss}",
    "{mean} ({sd})",
    "{median} ({p25}, {p75})"
  #  "{min}, {max}"
  ),
  percent="column",      # Include percentages for categorical variables by column options "column", "row" or "cell"
  missing="no"           # Exclude missing, options are "no", "ifany" and "always"
  ) |>
  gtsummary::modify_caption("Previous neck irradiation")
patient_characteristics

```

<!-- This is an example of how to have comments in Markdown files, if you want to comment out a section start with the -->
<!-- characters at the beinging of these lines and end with those at the end. -->
<!-- For options to inline_text() see https://www.danieldsjoberg.com/gtsummary/reference/inline_text.tbl_summary.html -->
Summary statistics are presented in @tbl-patient-characteristics-gtsummary. The mean and standard deviation of the age of
patients was `r gtsummary::inline_text(patient_characteristics, variable="age", level="Mean (SD)")`.
Height and/or weight were only available for
`r gtsummary::inline_text(patient_characteristics, variable="bmi", level="N")` patients, of these the mean and standard
deviation of BMI in patients was  `r gtsummary::inline_text(patient_characteristics, variable="bmi", level="Mean (SD)")`
whilst the median and inter-quartile range was
`r gtsummary::inline_text(patient_characteristics, variable="bmi", level="Median (IQR)")`.

<!-- The most common ASA Score was -->
<!-- Grade II (n = `r gtsummary::inline_text(patient_characteristics, variable="asa_score", level="II")`) -->
<!-- with Grade I being the second most common -->
<!-- (n = `r gtsummary::inline_text(patient_characteristics, variable="asa_score", level="I")`). -->

<!-- ```{r} -->
<!-- #| label: tbl-neck-irradiation-tableone -->
<!-- #| eval: false -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- #| tbl-caption: "Previous neck irradiation" -->
<!-- p_charac <- select(df, c( -->
<!--   "age", "bmi", "referral_source", "two_week_wait_referral", "smoking_status", -->
<!--   "previous_neck_irradiation", "asa_score" -->
<!-- )) -->
<!-- CreateTableOne( -->
<!--   data = p_charac, vars = c( -->
<!--     "age", "bmi", -->
<!--     "referral_source", -->
<!--     "two_week_wait_referral", -->
<!--     "smoking_status", -->
<!--     "previous_neck_irradiation", -->
<!--     "asa_score" -->
<!--   ), -->
<!--   factorVars = c("previous_neck_irradiation") -->
<!-- ) -->
<!-- ``` -->


```{r}
#| label: tbl-neck-symptoms-imaging-gtsummary
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Neck Symptoms and Imaging"
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
neck_symptoms_imaging <- df |>
  dplyr::ungroup() |>
  dplyr::select(c(
    "neck_symptoms", "incidental_imaging", "incidental_imaging_type", "abnormal_thyroid_function", "hyperthyroidism",
    "hypothyroidism", "no_symptoms", "neck_lump", "compressive_symptoms", "clinical_assessment", "retrosternal",
    "palpable_lymphadenopathy", "nodule_rapid_growth"
  )) |>
  gtsummary::tbl_summary(
    percent="column",      # Include percentages for categorical variables by column options "column", "row" or "cell"
    missing="no"           # Exclude missing, options are "no", "ifany" and "always"
  ) |>
  gtsummary::modify_caption("Neck Symptoms and Imaging")
neck_symptoms_imaging
```

Neck symptoms and images are summarised in @tbl-neck-symptoms-imaging-gtsummary Neck symptoms were recorded in
`r gtsummary::inline_text(neck_symptoms_imaging, variable="neck_symptoms")`
of cases and incidental imaging was performed in
`r gtsummary::inline_text(neck_symptoms_imaging, variable="incidental_imaging")`
cases of which  Ultrasound
(`r gtsummary::inline_text(neck_symptoms_imaging, variable="incidental_imaging_type", level="Ultrasound")`) and CT
(`r gtsummary::inline_text(neck_symptoms_imaging, variable="incidental_imaging_type", level="CT")`)
were the most common. Clinical assessment identified the most common nodule type to be single
(`r gtsummary::inline_text(neck_symptoms_imaging, variable="clinical_assessment", level="Single thyroid node")`).



<!-- ```{r} -->
<!-- #| label: tbl-neck-symptoms-imaging-tableone -->
<!-- #| eval: false -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- #| tbl-caption: "Neck Symptoms and Imaging" -->
<!-- c_charac <- select(df, c( -->
<!--   "neck_symptoms", "incidental_imaging", "incidental_imaging_type", -->
<!--   "abnormal_thyroid_function", -->
<!--   "hyperthyroidism", "hypothyroidism", "no_symptoms", "neck_lump", -->
<!--   "compressive_symptoms", "clinical_assessment", "retrosternal", -->
<!--   "palpable_lymphadenopathy", "nodule_rapid_growth" -->
<!-- )) -->
<!-- CreateTableOne( -->
<!--   data = c_charac, vars = c( -->
<!--     "neck_symptoms", "incidental_imaging", -->
<!--     "incidental_imaging_type", -->
<!--     "abnormal_thyroid_function", -->
<!--     "hyperthyroidism", "hypothyroidism", "no_symptoms", "neck_lump", -->
<!--     "compressive_symptoms", "clinical_assessment", "retrosternal", -->
<!--     "palpable_lymphadenopathy", "nodule_rapid_growth" -->
<!--   ), -->
<!--   factorVars = c( -->
<!--     "neck_symptoms","incidental_imaging", -->
<!--     "incidental_imaging_type", -->
<!--     "abnormal_thyroid_function", -->
<!--     "hyperthyroidism", "hypothyroidism", "no_symptoms", "neck_lump", -->
<!--     "compressive_symptoms", "clinical_assessment", "retrosternal", -->
<!--     "palpable_lymphadenopathy", "nodule_rapid_growth" -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->


```{r}
#| label: tbl-ultrasound-investigations-gtsummary
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Ultrasound investigations."
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
ultrasound_investigations <- df |>
  dplyr::ungroup() |>
  dplyr::select(
    c("nodule_ultrasound_description", "nodule_maxmimum_diameter_mm", "nodule_ultrasound_u_stage",
      "nodule_ultrasound_lymphadenopathy")) |>
  gtsummary::tbl_summary(
    percent="column",      # Include percentages for categorical variables by column options "column", "row" or "cell"
    missing="no"           # Exclude missing, options are "no", "ifany" and "always"
  ) |>
  gtsummary::modify_caption("Ultrasound Investigations.")
ultrasound_investigations
```

<!-- ```{r} -->
<!-- #| label: tbl-ultrasound-investigations-tableone -->
<!-- #| eval: false -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- #| tbl-caption: "Ultrasound investigations." -->
<!-- u_charac <- select(df, c( -->
<!--   "nodule_ultrasound_description", "nodule_maxmimum_diameter_mm", "nodule_ultrasound_u_stage", -->
<!--   "nodule_ultrasound_lymphadenopathy")) -->
<!-- CreateTableOne( -->
<!--   data = u_charac, vars = c( -->
<!--     "nodule_ultrasound_description", "nodule_maxmimum_diameter_mm", "nodule_ultrasound_u_stage", -->
<!--     "nodule_ultrasound_lymphadenopathy"), -->
<!--   factorVars = c( -->
<!--     "nodule_ultrasound_description", "nodule_ultrasound_u_stage", "nodule_ultrasound_lymphadenopathy") -->
<!-- ) -->
<!-- ``` -->

```{r}
#| label: tbl-fna-biopsy-investigations-gtsummary
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "FNA and Biopsy investigations."
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
fna_biopsy_investigations <- df |>
  dplyr::ungroup() |>
  dplyr::select(c("nodule_fna", "nodule_fna_result", "nodule_fna_thy", "core_biopsy")) |>
  gtsummary::tbl_summary(
    percent="column",      # Include percentages for categorical variables by column options "column", "row" or "cell"
    missing="no"           # Exclude missing, options are "no", "ifany" and "always"
  ) |>
  gtsummary::modify_caption("FNA and Bopisy investigations.")
fna_biopsy_investigations
```

<!-- ```{r} -->
<!-- #| label: tbl-fna-biopsy-investigations-tableone -->
<!-- #| eval: false -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- #| tbl-caption: "FNA and Biopsy investigations." -->
<!-- biopsy <- select(df, c( -->
<!--   "nodule_fna", "nodule_fna_result", "nodule_fna_thy", "core_biopsy")) -->
<!-- CreateTableOne( -->
<!--   data = biopsy, vars = c( -->
<!--     "nodule_fna", "nodule_fna_result", "nodule_fna_thy", -->
<!--     "core_biopsy"), -->
<!--   factorVars = c( -->
<!--     "nodule_fna", "nodule_fna_result", "nodule_fna_thy", "core_biopsy")) -->
<!-- ``` -->

```{r}
#| label: tbl-management-review-gtsummary
#| eval: false
#| echo: false
#| warning: false
#| tbl-caption: "Management and Review strategies."
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
management_review <- df |>
  dplyr::ungroup() |>
  dplyr::select(
    c("initial_management_decision", "routine_review_offered", "routine_review_offered_interval_weeks",
      "routine_review_ultrasound", "routine_review_fna", "routine_review_management_strategy_review",
      "routine_review_management_revised", "routine_review_management_revised_reason",
      "routine_review_confirm_management_plan", "routine_review_patient_signposting_information", "thyroid_surgery"
  )) |>
  gtsummary::tbl_summary(
    percent="column",      # Include percentages for categorical variables by column options "column", "row" or "cell"
    missing="no"           # Exclude missing, options are "no", "ifany" and "always"
  ) |>
  gtsummary::modify_caption("Management and Review strategies.")
management_review
```

<!-- ```{r} -->
<!-- #| label: tbl-management-review-tableone -->
<!-- #| eval: false -->
<!-- #| echo: false -->
<!-- #| warning: false -->
<!-- #| tbl-caption: "Management and Review strategies." -->
<!-- management <- select(df, c( -->
<!--   "initial_management_decision", "routine_review_offered", -->
<!--   "routine_review_offered_interval_weeks", "routine_review_ultrasound", -->
<!--   "routine_review_fna", "routine_review_management_strategy_review", -->
<!--   "routine_review_management_revised", "routine_review_management_revised_reason", -->
<!--   "routine_review_confirm_management_plan", "routine_review_patient_signposting_information", -->
<!--   "thyroid_surgery" -->
<!-- )) -->
<!-- tone <- CreateTableOne(data = management, vars = c( -->
<!--   "initial_management_decision", "routine_review_offered", -->
<!--   "routine_review_offered_interval_weeks", "routine_review_ultrasound", -->
<!--   "routine_review_fna", "routine_review_management_strategy_review", -->
<!--   "routine_review_management_revised", -->
<!--   "routine_review_management_revised_reason", -->
<!--   "routine_review_confirm_management_plan", -->
<!--   "routine_review_patient_signposting_information", -->
<!--   "thyroid_surgery" -->
<!-- )) -->
<!-- ``` -->


```{r}
#| label: tbl-final-pathology-gtsummary
#| eval: true
#| echo: false
#| warning: false
#| tbl-caption: "Final Pathology."
# IMPORTANT : For options to tbl_summary() see https://www.danieldsjoberg.com/gtsummary/reference/tbl_summary.html
final_pathology <- df |>
  dplyr::ungroup() |>
  dplyr::select(c("final_pathology", "nodule_fna", "age", "bmi", "smoking_status", "asa_score")) |>
    gtsummary::tbl_summary(by="final_pathology") |>
  add_p() |>
  gtsummary::modify_caption("Final Pathology.")
final_pathology
print(colnames(final_pathology))
```

### Patient characteristics

See Table for the clinical variables at presentation. Smokers and non-smokers represented 88 (33.7%) and 173 (66.3%)
respectively. Of note 312 (54.4%) of patients did not have smoking status recorded. Of those who had BMI recorded `r
mean(df$bmi, na.rm=TRUE)` (`r sd(df$bmi, na.rm=TRUE)`. A total of 460 patients had the American Society of
Anaesthesiologist (ASA) classifications recorded. The ASA classifications were as follows: ASA I, 186 (40.4%); ASA II,
201 (43.7%); ASA III, 70 (15.2%), ASA IV, 3 (0.7%).

### Modes of presentation

Table x shows the modes of presentations. A total of 392/573 (68.4%) of thyroid nodules presented with neck
symptoms. Other modes of presentation were as follows: biochemical hyperthyroidism, 26/573 (4.5%); biochemical
hypothyroidism, 4/573 (0.7%); symptoms of abnormal thyroid function, 22/573 (3.8%). Thyroid nodules were found
incidentally on imaging in 181/573 (31.6%) of patients; the imaging modality was recorded in 178 patients. Ultrasound
neck was the commonest modality for identification of incidental thyroid nodules 75/178 (42.1%). Other modality for
identification of incidental nodules were as follows: CT scan, 55/178 (30.9%); MRI, 24/178 (13.5%); nuclear medicine
scan 24/178, (13.5%).


### Clinical assessment
The following symptoms at presentation were recorded: neck lump, 364/573 (63.5%); compressive symptoms (breathing or
swallowing difficulty, or voice change), 87/573 (15.2%); symptoms of thyroid dysfunction, 34/573 (5.9%). A total of
136/573 (23.7%) of patients were considered asymptomatic at presentation.
Rapid perception of growth of the nodule was noted in 54/506 (10.7%) patients. Data on rapid perception of growth was
missing in 67 patients. A total of 519 patients had data provided on clinical palpable nodules. On clinical assessment,
265/519 (51.2%) of patients had a single palpable nodule. No nodule was palpable in 161 (31.0%) patients. Retrosternal
extension was suspected in 15/397 (3.8 %) patients. Data was missing on retrosternal extension in 176 patients. Thyroid
function test (TFT) was done within three months of presentation in 475/570 (83.3%) of patients.

### Ultrasound findings
Most patients 551/573 (96.2%) had an ultrasound scan either prior to or following initial assessment in secondary
care. Description of the thyroid nodule on ultrasound was available in 544 patients. A single thyroid nodule was seen in
204 (37.5%); more than one thyroid nodules were seen in 321 (59.0%) of cases. The median (IQR) maximum diameter of the
largest nodule was 23 mm (13-34). The BTA U classification for thyroid nodules was reported in 432 patients. The BTA U
classifications were recorded as follows: U1, 2 (0.5%); U2, 197 (45.6%); U3, 192 (44.4%); U4, 31 (7.2%); U5, 10
(2.3%). See table 7 for the distribution of BTA U classification in seven sites who reported in at least 30
patients. TIRADS reporting was use in 49/574 (8.6%) of patients. Associated abnormal cervical lymphadenopathy was
recorded in 26 patients (4.5%). Elastography was not used in any patient to assess the thyroid nodule.

### Other imaging
CT neck was performed to further evaluate the thyroid nodule in a total of 67/573 (11.7%) patients. Commonest reasons
for CT evaluation were retrosternal extension (18 patients) and tracheal compression (16 patients). Other reasons
documented were extrathyroidal extension/local invasion (7 patients) and extent of lymphadenopathy (7
patients). Furthermore, MRI neck was done in 25/573 (4.4%) patients; commonest reason was to evaluate for extent of
lymphadenopathy (7/25, 28%).

### Fine needle aspiration

FNAC FNAC was performed in a total of 377/573 patients (65.8%) of patients. of
these, The BTA Thy classification system was provided in the pathology report in 301 (79.8%) of patients who had
FNAC. The BTA Thy classifications of the nodules were as follows: Thy1, 66/301 (21.9%); Thy2, 113/301 (37.5%); Thy3a,
31/301 (10.3%); Thy3f, 63/301 (20.9%); Thy4, 14/301 (4.7%); Thy5, 14/301 (4.7%). The Thy classification system was not
reported in 76 (13.3%) patients. We examined four sites that had a least 30 patients with FNAC done, to see the
distribution of results. See Table 8, it highlight variability in the distribution of FNAC result. Core biopsy was done
in 8 patients

### Management
Data on the management of patients following initial investigations was availabe in 561 patients. At total of 417/561
(74.3%) patients were managed conservatively following initial assessment. Of the patients  were conservatively managed
following initial investigations (n=417), data on routine review was missing in 10 patients. Routine review was arranged
in 227/407 (55.8%) of patients; the median (IQR) time to review following initial assessment was 24 (12-24) weeks. The
following interventions after initial investigations were arranged: 135/561 (24.1%); radiofrequency ablation/ethanol
injection, 6/561 (1.1%); radioiodine, 3/561 (0.5%).  Of patients who had routine review following initial assessment, a
repeat ultrasound thyroid was done in 154/215 (71.6%) of patients; and 45/210 (21.4%) had a repeat FNAC. Furthermore,
management decision was changed in a total of 27 (13.4%) patients; all had surgical management. The others continued on
a conservative management strategy.

Of patients who had surgery, the surgery performed were as follows: isthmusectomy, 3 (2.0%); lobectomy, 113 (75.3%); and
total thyroidectomy, 34 (22.7%). The postoperative pathology were benign and malignant in 108 (58.7%) and 76 (41.3%)
respectively. The rates of malignant rates as per the different Thy classifications were: Thy1, 8/25  (32.0%); Thy2,
5/26  (19.2%); Thy3a, 9/21 (42.9%); Thy3f, 23/48 (47.9%); Thy4, 10/13  (76.9%); Thy 5, 14/14 (100%). he association
between clinical variables and thyroid cancer

### Predictors of malignancy
We evaluated potential predictors for malignancy. Palpable lymphadenopathy at presentation was significantly associated
with thyroid cancer in patients presenting with thyroid nodules. Other clinical factors were not clinically significant.

## Discussion
The multicentre observational study examined the presentation and initial management of thyroid nodules in secondary
care. The result has provided insights into characteritics of thyroid nodules presenting in the secondary care. Multiple
health visits, various investigations and, multiple specialties involved.

### Variability across sites in the multicentre study

There was significant variability in total number of patients contributed  by each site. The sites have different volume
of thyroid and endocrine surgery which may explain this variability. In addition, some sites were district general
hospitals which have a relatively lower volume of patients compared to the teaching hospital units that have a greater
volume of patients referred to them.  One potential reason for the variablity in the number of patients per centre could
be that not all consecutive patients during the study period were included. Given the restropective nature of the study,
and potential difficulty in accurately identifying all eligible patients, this is a distinct possibility. To improve
identification of all elibigle patients, we provided each centre with a flow chart and guidance to improve
identification of all suitable patients, thus minimising this risk. It was accepted that centres may have different
systems in place given the variability in methods of storing electronic records between hospitals.

### Referral for thyroid nodules

ENT clinics contributed the most number of patients (48.3%) to this study. General medical clinic and other specialty
contributed around 16% of patients to this study; this highlights the wide range of  specialties that initially assess
thyroid nodules in secondary care. Of note each site that were included in the study had thyroid surgeons or endocrine
physicians. It is thought that most thyroid nodules were referred from primary care, this was confirmed by our
results. However, a large proportion (26%) were inter specialty referral within secondary care. One reason for this is
the increasing identification of incidental thyroid nodule on imaging. We found around 52% of patients with thyroid
nodule were referred through the 2WW referral pathway. The national institute for health and care excellence (NICE) has
provided guidelines for clinicians on the recognition and referral of suspected thyroid cancer. They advised clinician
to consider referral within two weeks for thyroid cancer in patients with unexplained thyroid lump (Suspected cancer:
recognition and referral 2015). However, the BTA guidelines advised the two week rule to be applied only in patients
with additional suspicious features (i.e. unexplained hoarseness/voice change, thyroid nodule in a child, cervical
lymphadenopathy, a rapidly enlarging, painless, thyroid mass over a period of weeks) (4). The thyroid cancer guidelines
update group  recommends optimal care in patients with suspicious clinical features can be delivered  by using a target
of 4-week pathway from referral to assessment in clinic (4). Our findings suggest UK  practices still adopt  2WW
referral for the majority of patients. One explanation for this high proportion could be primary care physician
referring to NICE guidelines for unexplained thyroid lump. The BTA specialist guidelines may not be readily utilised by
primary care.

### Investigations for thyroid nodules

The majority of patients had thyroid function tests done before referral, this is in keeping with national
recommendations. This helps stratification of referral pathways; patients with suppressed TSH in association with a new
thyroid nodule are very unlikely to have thyroid cancer. Thus these patients are best referred to endocrinology; in
addition a non urgent referral is also advised @perros2014. Generally, most patients with a thyroid lump will undergo a thyroid
ultrasound, in keeping with national recommendations @mitchell2016 @perros2014. Of note, 3.8% of patients in our study did not
have an ultrasound at initial assessment. It is unclear why they did not. Given the rates of incidental nodules being
picked up in patients, it can be argued that not all patients with thyroid pathology require thyroid ultrasound. Around
60% of patients who had an ultrasound where found to have more than one thyroid nodules; this further emphasises how
common thyroid nodules are. The relatively high proportion of patients undergoing FNAC demonstrates the need for better
predictors to avoid unnecessary biopsies. The BTA U classification system outlined specific features for radiologist to
look out for in each category, with the view of reducing variability between assessors. However, varibility may still
exist between radiologists. Utilising machine learning approach may help in reducing variability and improve
stratification of nodules to benign and malignant @thomas2020. Risk stratification may be aided by the use of additional
predictive factors such as specific demographic and clinical features @sands2011. Ultrasound adjuncts such as elastrophy is
increasing being studied to help stratification of nodules into benign or malignant @RN3292. None of the centres
included in our study used this modality. It appears that it is not a technique adopted to the UK. It is thought most
patients undergoing thyroid surgery for thyroid nodules are young and fit patients with minimal comorbidity.Our results
showed that patients had ASA I-II status in only 80% of instances. Thus diagnostic thyroid surgery has the potential to
reduce the benefit risk ratio, particulary in the context were majority of patients have benign pathology. Incidental
nodules found on imaging represented around 30% of our study population. The definition of thyroid incidentaloma varies
in the literature; in this study  incidentaloma was defined as thyroid nodule found incidentally on imaging. Ultrasound
of the neck was the commonest modality for identification of incidentaloma, followed by CT scan. Nodules meeting
threshold for biopsy (U3 or above on ultrasound) are then referred for thyroid cytology. Studies have shown no
difference in the rates of cancer between clinically apparent nodule and incidental nodule @chooi2022. The increasing
incidence of thyroid cancer and the potential over diagnosis is largely attributed to identification of incidental
nodules @davies2006. However, if the incidentally detected nodule has similar risk of cancer to the clinically apparent
nodule, it will need to be investigated. This suggests that attempts to address thyroid cancer over diagnosis may be to
reduce need for imaging as opposed to limiting investigations of nodules detected on imaging. Furthermore a pragmatic
approach could be not commenting on incidental nodules less than 1 cm; as in the context of thyroid cancer, recent
guidelines that advised clinicians can consider a conservative approach. This is a challenging topic to address as
clinicians will be concerns about not excluding thyroid cancer and patients may have ongoing concerns even though the
literature will suggest most nodules are benign and subcentimetre thyroid nodules could be managed conservatively
@nice2022.

The terminology could be another point to consider, thyroid cancer in the context of micropapillary carcinoma
still creates anxieties for patients. autopsy study has show thyroid cancer in thyroid specimens that was never detected
clinically, these patients died from other causes(12). CT scan was rarely used in the evaluation for thyroid nodule; it
was reserved in cases of suspected retrosternal extension or extrathyroid extension which is in keeping with national
guidelines. Given most thyroid nodules are benign, we can assume a large proportion of nodules do not require FNAC. FNAC
is generally done in patients who meets threshold for suspicion of malignancy on ultrasound assessment. Limitations of
FNAC includes non diagnostic specimen and indeterminate results.  We found non diagnostic samples in 21% of
patients. Indeterminate results carry a variable rate of malignancy: 10-30% for Thy3a and 10 to 40% for Thy3f. This
varied across centres; likely attributed to variability in cytological assessments. In the setting of Thy3f, diagnostic
hemithyroidectomy is advised, and in Thy3a, repeat US and biopsy is typically done on follow-up. FNA was performed in
around 53% of patients in our study population. Of the patients who had FNAC, the Thy classification result were as
follows: Thy1 (21.9%), Thy2 (37.5%), Thy3a (10.3%), Thy3f (20.9%), Thy4 (4.7%), Thy5 (4.7%). This shows around 30% of
thyroid nodule nodules were indeterminate; 20% overall requiring diagnostic thyroidectomy. Looking at the four sites
that performed at least 30 FNAC in patients, there was significant variation in the frequency of indeterminate pathology
results. The proportion of Thy3a and Thy3f which had malignant pathology were 42.9% and 47.9% respectively. Thus, over
50% of the indeterminate nodules that undergo diagnostic surgery were benign on final pathology. This exposes a great
number of patents without thyroid cancer to morbidities associated with thyroid surgery. This highlights the need for
better discriminator in distinguishing between benign and malignant nodules.

### Molecular testing for thyroid nodules

Molecular testing is increasing being used in the evaluation thyroid nodules. Molecular markers have been developed to
improve decision making in patients with indeterminate results from FNA. Available molecular assays include Affirma®
GSC; ThyGenX/ThyraMIR® ThyraGeNEXT/ThyraMIR®; and Thyroseq v3®(@nylen2020). Molecular markers identify genetic mutations
that are associated with malignancy to further aid stratification of indeterminate nodules. Most patients with
indeterminate nodules have a thyroid lobectomy for definitive diagnosis; a negative molecular test result that has a
high negative predictive value has the potential to reduce the need for diagnostic surgery. The accuracy of the
molecular tests is however affected by the prevalence of malignancy. Thus, a lower prevalence of malignancy will result
in a lower positive predictive value, and a higher negative predictive value @morris2020. The utility of these molecular
markers in patients with indeterminate FNA result is under debate. This is because a large proportion of patients with
indeterminate nodules on FNA eventually are found to have benign disease. In addition, malignancy in this group
represents well differentiated thyroid cancer with an excellent prognosis (@morris2020). These molecular tests have a
low PPV (20-40%) for malignancy in patients with indeterminate nodules (@morris2020). This is because the probability of
an indeterminate harbouring malignancy is inherently low (@haugen2016). In addition, the genetic alterations identified
with the assays are not solely unique to carcinoma; these can also be seen in adenomas (@morris2020). Caution has been
advised with interpretation of the tests (16). These tests are not available for routine use in the UK.

### Follow-up for patients with thyroid nodules.

In cases where no intervention was planned following initial assessment, 55% of patients were offered routine review in
a median period of 24 weeks. This highlights the increase workload associated with the management of thyroid nodules;
this workload created by the uncertainty in FNAC results. Of the patients who had routine review, 15% had a change in
the initial management strategy.

### Potential risk factors for thyroid cancer evaluated in the multicentre study.

On further analyses of the potential associations between risk factors and thyroid cancer, the following factors were not
significantly associated with thyroid cancer in patients presenting with thyroid nodules: age, smoking, BMI, previous
neck radiation, rapid growth and nodule size. As expected, associated palpable lymphadenopathy was significantly
associated with thyroid cancer. Exposure to ionising radiation (especially during childhood) is an established risk
factor for thyroid cancer @iglesias2017. A history of radiation exposure has also been shown to increase risk of thyroid cancer
in patients with thyroid nodules @sun2023.  This study found no significant difference in the rates of thyroid cancer
between patients with and without history of previous neck radiation; but the results may be subject to recall bias or
small numbers resulting in a type II error. Thyroid incidentaloma are thought to carry similar risk of thyroid cancer to
nodules detected on clinical examination @russ2014. Our findings found no significant differences in the rates of
thyroid cancer between incidental nodules on imaging and other nodules detected clinically. This supports the increasing
referral of thyroid nodules detected incidentally. Although these nodules could represent low risk papillary thyroid
cancer, they may never become clinically apparent during the course of the patient’s life. Perception of rapid growth is
considered a concerning feature in a patient presenting with thyroid nodule; as this can indicate poorly differentiated
or anaplastic thyroid cancer. This study did not demonstrate any association between perception of rapid growth and
thyroid cancer.

### Improving prediction of thyroid cancer in patients with thyroid nodules

There is a need for improved differentiation between benign and malignant nodules. The traditional practice of medicine
involves self-learning, acquisition of knowledge and experience to make an informed decision @chen2019. This process is
vital given the highly varied clinical environment, where decision making is based not only science but also the
environment and psychosocial factors. Advances in our knowledge and understanding has created large data sets of
information including a myriad of variables. There are complex relationships between variables that are poorly
understood but these influence outcomes @sidey-gibbons2019.  Machine learning to improve the understanding of these
relationships is gaining popularity in many areas of medicine @chen2019. Machine learning allows quick processing of
large, complex data sets to estimate probability of an outcome. These techniques often have an iterative process that
improve the precision over time with the addition of more information/data set @sidey-gibbons2019. Furthermore,
machine learning allows “personalised” medicine, whereby outcome data is generated based on an individual’s
characteristics rather than fixed group probabilities @schork2019. This property of machine learning has been adopted
in medical practice to calculate individual risk of an outcome and direct informed decision making on personalised
treatment @schork2019. The use of statistical models using several variables to predict cancer risk in patients with
thyroid nodules has been explored. Nixon et al. developed a nomogram to predict thyroid cancer in patients presenting
with thyroid nodules @nixon2010. They included all patients who had ultrasound, FNA and resection of thyroid nodule over
a one-year period (158 patients). Factors included in their nomogram included thyroid stimulating hormone; ultrasound
and cytological features. This tool is based on a fixed statistical model with no process of iteration. The utility of
this tool is not established and had not been adopted to routine clinical practice. Artificial intelligence (AI) has
been evaluated to reduce subjectivity in ultrasound reporting. One study examined ultrasound images of thyroid nodules
in 482 patients who had resection on the thyroid gland with definitive diagnosis of benign or malignant disease
@thomas2020. These images were then used to create an AI algorithm for prediction of malignant disease in patients who
had ultrasound of thyroid nodules. They reported a NPV of 93%, and a PPV of 81.5%. This was comparable to other reported
ultrasound systems. Other AI models have also shown comparative performance to radiologist assessment @peng2021. The
role of AI in improving diagnosis of thyroid cytology specimen has been explored using various machine learning
techniques @kezlarian2020. Some authors use the image from the cytology slides, others used data derived from the
images. Varying accuracy rates have been reported. These techniques are currently under investigation for further
refinement. In patients with indeterminate nodules, AI techniques have been used to stratify in high versus low risk of
malignancy (@elliot_range2020). One study evaluated stratification of 371 patients with indeterminate nodules using
cytological variable found that AI techniques was better than standard criteria used by pathologist. The AI classified
the indeterminate nodules into a high-risk group (33% risk of malignancy) and a low-risk group (3% risk of malignancy)
@ippolito2004. There is a however a lack of studies that combine data from clinical features, ultrasound and
cytological variables in a model that may have the ability to improve the prediction of cancer than any one set of
variables.

## Conclusion

Around 60% of patients who underwent diagnostic thyroidectomy for thyroid nodules were found to have benign
pathology. This highlights the need for better preoperative descriminators between benign and malignant thyroid nodules,
to avoid unncessary diagnostic surgery and its associated morbidity.

## Notes from Neil

### Code Chunks

Give them all unique names.

### Use more tables and figures

Long chunks of text with number and percentages are tedious to read at best, more likely they will be ignored. Far
better to use tables and figures and draw readers attentiont to a few key points in these. Some examples have been
added, the tables were generated using the
[`{gtsummary}`](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html) package.

Where you do want to refer to numbers from tables in the text first generate the table as a `gtsummary::tbl_summary()`
object, you can then use the [inline text](https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html) functions
of this package to extract data and include it in the text where required.

### Citation Style

The citation style/format can be changed, see [Citation
Style](https://quarto.org/docs/authoring/footnotes-and-citations.html#sec-citations-style).
